```{r setup, include=F, cache=F}
library(tidyverse)
library(readxl)
library(ggrepel)
library(knitr)
library(glue)
library(patchwork)
opts_chunk$set(echo=F, cache=T, message=F, use_shortcode=T)
```

```{r file="ipeds.R"}
```

```{r download}
data_dir <- "ipeds"
if (!dir.exists(data_dir)) {
    message("Creating data directory ", data_dir)
    dir.create(data_dir)
}

years <- 2013:2020

eap_files <- get_ipeds("EAP", years, "")
```

```{r load}
eap <- load_series(eap_files)
```

EAP coding is annoying because many categories are merely supercategories of others. The exception is the most general level "All staff" which sometimes is not exhausted by apparent subcategories.

EAPCAT is a redundant variable:

        EAPCAT = OCCUPCAT . FACSTAT[0]

with the 0 padding only used for `FACSTAT = 0`. Verification:

```{r}
eap %>% mutate(FACSTAT=ifelse(FACSTAT == 0, "00", FACSTAT)) %>%
    mutate(ck=EAPCAT == str_c(OCCUPCAT, FACSTAT)) %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```

The count variables `EAP*` are also redundant:

```{r}
eap %>% summarize(ck=all(EAPTOT == EAPTYP + EAPMED, na.rm=T)) %>%
    pull(ck)
eap %>% summarize(ck=all(EAPFT == EAPFTTYP + EAPFTMED, na.rm=T)) %>%
    pull(ck)
eap %>% summarize(ck=all(EAPPT == EAPPTTYP + EAPPTMED, na.rm=T)) %>%
    pull(ck)
eap %>% summarize(ck=all(EAPTYP == EAPPTTYP + EAPFTTYP, na.rm=T)) %>%
    pull(ck)
eap %>% summarize(ck=all(EAPMED == EAPPTMED + EAPFTMED, na.rm=T)) %>%
    pull(ck)
```

To check identities in the coding of OCCUPCAT and FACSTAT, the following function can be used

```{r}
verify_identity <- function (x, var, partcodes, totcode) {
    x %>% filter({{ var }} %in% c(partcodes, totcode)) %>%
        filter(any({{ var }} %in% partcodes)) %>%
        mutate(tot=ifelse({{ var }} == {{ totcode }}, 1 , -1)) %>%
        summarize(ck=sum(EAPTOT * tot) == 0) %>%
        summarize(ck=all(ck))
}
```

# OCCUPCAT identities


        # 200 = 210 + 220 + 230

Instr, research, public service: holds

```{r}
eap %>% group_by(year, UNITID, FACSTAT) %>%
    verify_identity(OCCUPCAT, c(210, 220, 230), 200) %>%
    ungroup() %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```

        # 210 = 211 + 215

Instructional staff identity: holds except for one year of the Sacramento Ultrasound Institute

```{r}
eap %>% group_by(year, UNITID, FACSTAT) %>%
    verify_identity(OCCUPCAT, c(211, 215), 210) %>% filter(!ck)
```

        # 211 = 212 + 213 + 214

Instruction cr/non-cr identity: holds

```{r}
eap %>% group_by(year, UNITID, FACSTAT) %>%
    verify_identity(OCCUPCAT, 212:214, 211) %>%
    ungroup() %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```

        # 250 = 260 + 270

Librarians, Archivists, etc:

```{r}
eap %>% group_by(year, UNITID, FACSTAT) %>%
    verify_identity(OCCUPCAT, c(260, 270), 250) %>%
    ungroup() %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```

        # 260 = 261 + 262 + 263 + 264

Librarian subcategories:

```{r}
eap %>% group_by(year, UNITID, FACSTAT) %>%
    verify_identity(OCCUPCAT, 261:264, 260) %>%
    ungroup() %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```

        # 400 = 4[1..8]0 | up through 2015

Grad subcategories. In 2013–2015, there was a more refined taxonomy of grads and the identity is

```{r}
eap %>% filter(year %in% 2013:2015) %>%
    group_by(year, UNITID, FACSTAT) %>%
    verify_identity(OCCUPCAT, 10 * (41:48), 400) %>%
    ungroup() %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```

After that, we have

        # 400 = 410 + 420 + 490 | from 2016 on

```{r}
eap %>% filter(year >= 2016) %>%
    group_by(year, UNITID, FACSTAT) %>%
    verify_identity(OCCUPCAT, c(410, 420, 490), 400) %>%
    ungroup() %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```

# FACSTAT identities

        0 = 10 + 50 + 60 | up through 2016
        0 = 10 + 50 | unless OCCUPCAT=100


All staff either do or do not have faculty status, you'd think. Up through 2016 there is an NA category, FACSTAT = 60 ("non-degree granting institution"), which exhausts the possibilities. But after 2016, in the "All staff" OCCUPCAT=100 category, this NA category is gone and the dichotomy does not hold.


```{r}
eap %>% 
    filter(OCCUPCAT != 100) %>%
    group_by(year, UNITID, OCCUPCAT) %>%
    verify_identity(FACSTAT, c(10, 50), 0) %>%
    ungroup() %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```


        10 = 20 + 30 + 40

Faculty categories.

```{r}
eap %>% 
    group_by(year, UNITID, OCCUPCAT) %>%
    verify_identity(FACSTAT, c(20, 30, 40), 10) %>%
    ungroup() %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```


        40 = 41 + 42 + 43

NTT categories

```{r}
eap %>% 
    group_by(year, UNITID, OCCUPCAT) %>%
    verify_identity(FACSTAT, 41:43, 40) %>%
    ungroup() %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```

        41 = 44 + 45

NTT multiyear and indefinite breakdown

```{r}
eap %>% 
    group_by(year, UNITID, OCCUPCAT) %>%
    verify_identity(FACSTAT, 44:45, 41) %>%
    ungroup() %>%
    summarize(ck=all(ck)) %>%
    pull(ck)
```

        # 100 != 200 + 250 + 3[0..9]0 + 400

You would think the list of staff categories is meant to be exhaustive, but it is not. This equation is only true about half the time (counting each year/UNITID/OCCUPCAT separately).

```{r}
eap %>%
    group_by(UNITID, FACSTAT) %>%
    verify_identity(OCCUPCAT, c(200, 250, 10 * (30:39), 400), 100) %>%
    summarize(sum(ck), n())
```
